Optimal forward modeling of gravitational fields using tesseroids with adaptive discretization

### 2014-09-01 body.tex
Improvements to methods and finished algorithm

The accuracy of the integration
depends on the number of nodes, or GLQ order,
used in the discretization.
However, \citet{Ku1977} showed
that it also depends on the ratio between
the distance to the computation point
and the distance between adjacent nodes.
Fig.~\ref{fig:sample}
illustrates this effect
on the $g_{xy}$ gravity gradient component.
To increase the accuracy of the integration,
one could increase the number of nodes,
as in Fig.~\ref{fig:sample}c.
Likewise,
one could keep the number of nodes fixed
and instead divide the tesseroid into smaller ones.
This would increase the total number of nodes
and effectively decrease
the distance between adjacent nodes.
The latter is the approach
taken by \citet{Li2011}.
Here, we will follow the same approach
and make some improvements to the algorithm.
We will also attempt to quantify
the error committed in the numerical integration.

@@

We have developed a modified version
of the recursive division algorithm proposed by \citet{Li2011}.
One of the modifications made is that
we fix the number of GLQ nodes as
$N^\lambda=N^\phi=N^r=2$.
In this way, the distances between adjacent nodes
are approximately the dimensions of the tesseroid
\citep[see][]{Wild-Pfeiffer2008}.
Thus, we may use the tesseroid dimensions
as proxies for the distance between nodes.
We define the dimensions of a tesseroid
in the radial, latitudinal, longitudinal directions, respectively,
as

@@

\noindent
where $R=6378137\ m$ is the mean Earth radius
and the longitudes and latitudes are given in radians.

We define the distance $d$ as
the Cartesian distance between
the computation point $P$
and the center of the top face of the tesseroid
(Fig.~\ref{fig:ratio}a).

DESCRIBE ALGORITHM FIRST.

Thus, the rule-of-thumb of \citet{Ku1977}
can be interpreted as

\begin{equation}
    \frac{d}{L_\zeta} < r,
    \label{eq:ratio}
\end{equation}

\noindent
where
$L_\zeta$, $\zeta \in \{r, \phi, \lambda\}$,
is one of the dimensions of the tesseroid
and
$r$ is the maximum allowed ratio between distance and size.
For \citet{Ku1977}, this ratio should be equal to one.
The dimensions of the tesseroid are

In case the tesseroid is divided,
the inequality in equation \ref{eq:ratio}
is checked for each of the eight parts.
This process is repeated recursively
until no divisions are required,
ensuring that equation \ref{eq:ratio}
holds for all tesseroids created.
The gravitational effect
of the original tesseroid
is then calculated as
the sum of the effect
of the smaller tesseroids.

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using $R$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
    }
    \label{fig:ratio}
\end{figure}

The rule-of-thumb of \citet{Ku1977}
used by \citet{Li2011}
is equivalent to setting $R=1$.
Fig.~\ref{fig:ratio}
illustrates the subdivision of a tesseroid
(Fig.~\ref{fig:ratio}a)
for computation point $P$
using increasing values of $R$.
Note that increasing $R$
results is a finer division of the tesseroid
around the computation point
and thus a more accurate GLQ integration.
However, higher values of $R$
also result in
larger computation times.
For example,
the tesseroid in Fig.~\ref{fig:ratio}a
is replaced by
50 tesseroids for $R=0.5$ (Fig.~\ref{fig:ratio}b),
344 tesseroids for $R=1$ (Fig.~\ref{fig:ratio}c),
and 2472 tesseroids for  $R=2$ (Fig.~\ref{fig:ratio}d).

Clearly,
there is need to determine
an optimal value of $R$
that yields the desired accuracy
in the GLQ integration.
In the following section,
we compare the computed gravitational effects
of a tesseroid model
with that of a spherical half-shell
for which exist analytical expressions.

\subsection{Determination of $r$ by comparison with a spherical shell}

>>>>

The accuracy of the integration
depends on the number of nodes, or GLQ order,
used in the discretization.
However, \citet{Ku1977} showed
that it also depends on the ratio between
the distance to the computation point
and the distance between adjacent nodes.
Fig.~\ref{fig:sample}
illustrates this effect
on the $g_{xy}$ gravity gradient component.
To increase the accuracy of the integration,
one could increase the number of nodes,
as in Fig.~\ref{fig:sample}c.
Likewise,
one could keep the number of nodes fixed
and instead divide the tesseroid into smaller ones.
This would increase the total number of nodes
and effectively decrease
the distance between adjacent nodes.
The latter is the approach
taken by \citet{Li2011}.
Here, we will follow the same approach
and make some improvements to the algorithm.

@@

We start by fixing the number of GLQ nodes to
$N^\lambda=N^\phi=N^r=2$.
In this way, the distances between adjacent nodes
are approximately the dimensions of the tesseroid
\citep[see][]{Wild-Pfeiffer2008}.
Thus, we may use the tesseroid dimensions
as proxies for the distance between nodes.
We define the dimensions of a tesseroid
in the radial, latitudinal, longitudinal directions, respectively,
as

@@

\noindent
where $R=6,378,137\ m$ is the mean Earth radius.
The longitudes and latitudes are given in radians.

We write the condition for an accurate GLQ integration
described by \citet{Ku1977}
as a function of
the average of the dimensions of the tesseroid, $L$,

\begin{equation}
    \frac{d}{L} > 1,
\end{equation}

\noindent
where $d$ is the Cartesian distance
between the computation point $P$
and the center of the top face of the tesseroid
(Fig.~\ref{fig:ratio}a).
In a more general form, the condition can be expressed as

\begin{equation}
    \frac{d}{L} > r,
    \label{eq:ratio}
\end{equation}

\noindent
for a positive non-zero scalar $r$,
henceforth referred to as
the ``distance-size ratio''.

To guarantee that equation~\ref{eq:ratio} is satisfied,
one could divide the tesseroid in half for each of the three dimensions.
The division would be repeated recursively
for each of the eight smaller tesseroids
until equation~\ref{eq:ratio} is satisfied for all smaller tesseroids.
Then, the effect of the original tesseroid can be calculated
as the sum of the
effects of the smaller tesseroids,
each of which is calculated using a second-order GLQ.

However, this would result in unnecessary divisions,
specially if one dimension of the tesseroid is much smaller than the others.
A more efficient approach is to check if the condition is met for each of the
three dimensions of the tesseroid ($L_r$, $L_\phi$, and $L_\lambda$).
We only divide the tesseroid in half along dimensions for which the condition
is not met.

This algorithm can be implemented
in the form of a recursive function
as follows:

\begin{verbatim}
function gravity(P, model):
  result = 0
  for each tesseroid in model:
    d = distance from P to the top of tesseroid
    if d < r*L_lon:
      divide_lon = true
    if d < r*L_lat:
      divide_lat = true
    if d < r*L_r:
      divide_r = true
    if divide_lon or divide_lat or divide_r:
      newmodel = (divide tesseroid in the
                  required dimensions)
      result = result + gravity(P, newmodel)
    else
      result = result + effect of tesseroid
  return result
\end{verbatim}

The distance-size ratio $r$ controls
how far away $P$ has to be
compared to the dimensions of the tesseroid.
A ratio $r=2$ means that
the recursive division algorithm will ensure that
tesseroids are smaller than
half the distance to $P$.
Thus, the ratio determines how many divisions are required.
Fig.~\ref{fig:ratio}
illustrates the recursive division of a tesseroid
(Fig.~\ref{fig:ratio}a)
using increasing values of $r$.
The tesseroid is divided into
$28$ tesseroids for $r=1$ (Fig.~\ref{fig:ratio}b),
$146$ tesseroids for $r=2$ (Fig.~\ref{fig:ratio}c),
and $3106$ tesseroids for $r=6$ (Fig.~\ref{fig:ratio}d).

Because the number of tesseroids increases with $r$,
so does the computation time.
Hence, there is need to determine
an optimal value of the distance-size ratio $r$
that yields an acceptable trade-off between
accuracy and computation time.
In the following section,
we describe a methodology
to quantify the error of the GLQ integration
as a function of the distance-size ratio $r$.
These results can be used
to determine optimal values of $r$
for the gravitational potential,
gravitational attraction,
and gravity gradient tensor.

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $r$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $r$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:ratio}
\end{figure}

\subsection{Comparison with a spherical shell}
